# {{town_name}}

```{r}
#| label: process-age-{{town_name}}
#| include: false

# 1. Define the variables and give them readable names
age_vars <- c(
  "0 to 4"   = "S0101_C01_002",
  "5 to 9"   = "S0101_C01_003",
  "10 to 14" = "S0101_C01_004",
  "15 to 19" = "S0101_C01_005",
  "20 to 24" = "S0101_C01_006",
  "25 to 29" = "S0101_C01_007",
  "30 to 34" = "S0101_C01_008",
  "35 to 39" = "S0101_C01_009",
  "40 to 44" = "S0101_C01_010",
  "45 to 49" = "S0101_C01_011",
  "50 to 54" = "S0101_C01_012",
  "55 to 59" = "S0101_C01_013",
  "60 to 64" = "S0101_C01_014",
  "65 to 69" = "S0101_C01_015",
  "70 to 74" = "S0101_C01_016",
  "75 to 79" = "S0101_C01_017",
  "80 to 84" = "S0101_C01_018",
  "85+"      = "S0101_C01_019"
)

# 2. Fetch the data for all of MA, then filter to the specific town
town_age_raw <- get_acs(
  geography = "county subdivision",
  variables = age_vars,
  state = "MA",
  year = current_acs_year,
  survey = "acs5"
) |> 
  filter(str_detect(NAME, "^{{town_name}}"))

# 3. Roll up into the new, tighter brackets
town_age_summary <- town_age_raw |> 
  mutate(
    Bracket = case_when(
      variable %in% c("0 to 4", "5 to 9", "10 to 14", "15 to 19") ~ "Under 20 (Youth)",
      variable %in% c("20 to 24", "25 to 29", "30 to 34") ~ "20 to 34",
      variable %in% c("35 to 39", "40 to 44", "45 to 49", "50 to 54", "55 to 59", "60 to 64") ~ "35 to 64",
      TRUE ~ "65+ (Older Adults)"
    ),
    Bracket = factor(Bracket, levels = c("Under 20 (Youth)", "20 to 34", "35 to 64", "65+ (Older Adults)"))
  ) |> 
  group_by(Bracket) |> 
  summarize(
    Estimate = sum(estimate, na.rm = TRUE),
    MOE = sqrt(sum(moe^2, na.rm = TRUE)) 
  )
```

### Age Distribution
The following chart highlights the age distribution for **{{town_name}}**, breaking out youth and older adult populations.

```{r}
#| label: plot-age-{{town_name}}
#| echo: false
#| fig-width: 6
#| fig-height: 2

ggplot(town_age_summary, aes(x = Bracket, y = Estimate)) +
  geom_col(fill = "#2c3e50", width = 0.7) +
  geom_text(aes(label = scales::comma(Estimate)), vjust = 1.5, size = 3.5, color = "white") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = paste("Age Demographics:", "{{town_name}}"),
    x = NULL,
    y = "Population"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 11),
    axis.text.x = element_text(size = 10),
    panel.grid.major.x = element_blank()
  )
```

### Language & Proficiency

```{r}
#| label: process-language-{{town_name}}
#| include: false

# 1. Limited English Households (S1602)
lep_vars <- c(
  "Total Households" = "S1602_C01_001",
  "LEP Households" = "S1602_C03_001"
)

town_lep_raw <- get_acs(
  geography = "county subdivision",
  variables = lep_vars,
  state = "MA",
  year = current_acs_year,
  survey = "acs5"
) |> 
  filter(str_detect(NAME, "^{{town_name}}"))

# Calculate total LEP households and percentage
lep_total <- town_lep_raw |> 
  filter(variable == "LEP Households") |> 
  pull(estimate)

hh_total <- town_lep_raw |> 
  filter(variable == "Total Households") |> 
  pull(estimate)

lep_percent <- if_else(hh_total > 0, round((lep_total / hh_total) * 100, 1), 0)

# 2. Top 5 Languages (C16001)
lang_vars <- c(
  "Spanish" = "C16001_003",
  "French, Haitian, or Cajun" = "C16001_006",
  "German or other West Germanic" = "C16001_009",
  "Russian, Polish, or other Slavic" = "C16001_012",
  "Other Indo-European" = "C16001_015",
  "Korean" = "C16001_018",
  "Chinese (incl. Mandarin, Cantonese)" = "C16001_021",
  "Vietnamese" = "C16001_024",
  "Tagalog (incl. Filipino)" = "C16001_027",
  "Other Asian and Pacific Island" = "C16001_030",
  "Arabic" = "C16001_033",
  "Other Languages" = "C16001_036"
)

town_lang_raw <- get_acs(
  geography = "county subdivision",
  variables = lang_vars,
  state = "MA",
  year = current_acs_year
) |> 
  filter(str_detect(NAME, "^{{town_name}}"))

# Sort and keep the top 5 spoken languages
top_5_languages <- town_lang_raw |> 
  arrange(desc(estimate)) |> 
  filter(estimate > 0) |>  # Exclude languages with zero speakers
  slice_head(n = 5) 
```

In **{{town_name}}**, there are **`r scales::comma(lep_total)`** households considered to have Limited English Proficiency (LEP), representing **`r lep_percent`%** of total households. 

The top languages spoken at home (other than English) are:

```{r}
#| label: table-lang-{{town_name}}
#| echo: false

# Generate a simple, clean table for the Word doc
top_5_languages |> 
  select(Language = variable, Estimate = estimate) |> 
  gt() |> 
  cols_label(
    Language = md("**Language**"),
    Estimate = md("**Estimated Speakers**")
  ) |> 
  fmt_number(
    columns = Estimate,
    decimals = 0,
    use_seps = TRUE # Ensures numbers like 1200 render as 1,200
  ) |> 
  cols_align(
    align = "left",
    columns = Language
  ) |> 
  cols_align(
    align = "right",
    columns = Estimate
  ) |> 
  opt_stylize(style = 1, color = "gray") |> 
  tab_options(table.align = "left")
```


### Transportation Profile

```{r}
#| label: process-transportation-{{town_name}}
#| include: false

# 1. Vehicles Available (B08201)
veh_vars <- c(
 "Total Households" = "B08201_001",
  "No vehicle" = "B08201_002", 
 "1 vehicle" = "B08201_003",
 "2 vehicles" = "B08201_004",
 "3 vehicles" = "B08201_005", 
 "4 or more vehicles" = "B08201_006"
)

town_veh_raw <- get_acs(
  geography = "county subdivision",
  variables = veh_vars,
  state = "MA",
  year = current_acs_year
) |> 
  filter(str_detect(NAME, "^{{town_name}}"))

# Calculate Zero-Vehicle Households
zero_veh_total <- town_veh_raw |> 
  filter(variable == "No vehicle") |> 
  pull(estimate)

total_hh_veh <- town_veh_raw |> 
  filter(variable == "Total Households") |> 
  pull(estimate)

zero_veh_percent <- if_else(total_hh_veh > 0, round((zero_veh_total / total_hh_veh) * 100, 1), 0)

# 2. Means of Transportation to Work (B08301)
transpo_vars <- c(
  "Total Workers" = "B08301_001",
  "Drove alone" = "B08301_003", 
  "Carpooled" ="B08301_004", 
  "Public transportation" = "B08301_010", 
  "Taxi or ride-hailing services" = "B08301_016",
  "Motorcycle" = "B08301_017", 
  "Bicycle" = "B08301_018", 
  "Walked" = "B08301_019", 
  "Other means" = "B08301_020", 
  "Worked from home" = "B08301_021"
)

town_transpo_raw <- get_acs(
  geography = "county subdivision",
  variables = transpo_vars,
  state = "MA",
  year = current_acs_year
) |> 
  filter(str_detect(NAME, "^{{town_name}}"))

# Clean up for the chart
town_transpo_clean <- town_transpo_raw |> 
  filter(variable != "Total Workers", estimate > 0) |> 
  mutate(variable = reorder(variable, estimate))
```

In **{{town_name}}**, there are **`r scales::comma(zero_veh_total)`** households with no vehicles available, which represents **`r zero_veh_percent`%** of all households in the community.

**Means of Transportation to Work**

```{r}
#| label: plot-transpo-{{town_name}}
#| echo: false
#| fig-width: 6
#| fig-height: 3.5

ggplot(town_transpo_clean, aes(x = estimate, y = variable)) +
  geom_col(fill = "#d35400", width = 0.7) +
  geom_text(aes(label = scales::comma(estimate)), hjust = -0.1, size = 3.5) +
  scale_x_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.2))) +
  labs(
    title = paste("Commuting Patterns:", "{{town_name}}"),
    x = "Number of Workers",
    y = NULL
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 11),
    panel.grid.major.y = element_blank()
  )
```

```{=openxml}
<w:p><w:r><w:br w:type="page"/></w:r></w:p>
```