# {{town_name}}

```{=typst}
#columns(2, gutter: 4%)[
```

```{r}
#| label: process-age-{{town_name}}
#| include: false

age_vars <- c(
  "0 to 4"   = "S0101_C01_002",
  "5 to 9"   = "S0101_C01_003",
  "10 to 14" = "S0101_C01_004",
  "15 to 19" = "S0101_C01_005",
  "20 to 24" = "S0101_C01_006",
  "25 to 29" = "S0101_C01_007",
  "30 to 34" = "S0101_C01_008",
  "35 to 39" = "S0101_C01_009",
  "40 to 44" = "S0101_C01_010",
  "45 to 49" = "S0101_C01_011",
  "50 to 54" = "S0101_C01_012",
  "55 to 59" = "S0101_C01_013",
  "60 to 64" = "S0101_C01_014",
  "65 to 69" = "S0101_C01_015",
  "70 to 74" = "S0101_C01_016",
  "75 to 79" = "S0101_C01_017",
  "80 to 84" = "S0101_C01_018",
  "85+"      = "S0101_C01_019"
)

town_age_raw <- get_acs(
  geography = "county subdivision",
  variables = age_vars,
  state = "MA",
  year = current_acs_year,
  survey = "acs5"
) |> 
  filter(str_detect(NAME, "^{{town_name}}"))

town_age_summary <- town_age_raw |> 
  mutate(
    Bracket = case_when(
      variable %in% c("0 to 4", "5 to 9", "10 to 14", "15 to 19")                          ~ "Under 20 (Youth)",
      variable %in% c("20 to 24", "25 to 29", "30 to 34")                                   ~ "20 to 34",
      variable %in% c("35 to 39", "40 to 44", "45 to 49", "50 to 54", "55 to 59", "60 to 64") ~ "35 to 64",
      TRUE                                                                                    ~ "65+ (Older Adults)"
    ),
    Bracket = factor(Bracket, levels = c("Under 20 (Youth)", "20 to 34", "35 to 64", "65+ (Older Adults)"))
  ) |> 
  group_by(Bracket) |> 
  summarize(
    Estimate = sum(estimate, na.rm = TRUE),
    MOE = sqrt(sum(moe^2, na.rm = TRUE))
  ) |>
  mutate(
    Share = round((Estimate / sum(Estimate)) * 100, 1)
  )
```
```{r}
#| label: prose-age-{{town_name}}
#| include: false

youth_pct <- town_age_summary |>
  filter(Bracket == "Under 20 (Youth)") |>
  pull(Share)

older_adult_count <- town_age_summary |>
  filter(Bracket == "65+ (Older Adults)") |>
  pull(Estimate)

older_adult_pct <- town_age_summary |>
  filter(Bracket == "65+ (Older Adults)") |>
  pull(Share)
```

####
### Age Distribution

In **{{town_name}}**, youth under 20 make up **`r youth_pct`%** of the population, while **`r scales::comma(older_adult_count)`** residents (**`r older_adult_pct`%**) are aged 65 or older.
```{r}
#| label: plot-age-{{town_name}}
#| echo: false
#| fig-height: 1.5

ggplot(town_age_summary, aes(x = Bracket, y = Estimate)) +
  geom_col(fill = "#A00D03", width = 0.7) +
  geom_text(
    aes(label = paste0(scales::comma(Estimate), "\n(", Share, "%)")),
    vjust = -0.3,
    size = 2.2,
    lineheight = 0.9
  ) +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.35))
  ) +
  labs(
    x = NULL,
    y = "Population"
  )
```

### Language & Proficiency

```{r}
#| label: process-language-{{town_name}}
#| include: false

# 1. Limited English Households (S1602)
lep_vars <- c(
  "Total Households" = "S1602_C01_001",
  "LEP Households" = "S1602_C03_001"
)

town_lep_raw <- get_acs(
  geography = "county subdivision",
  variables = lep_vars,
  state = "MA",
  year = current_acs_year,
  survey = "acs5"
) |> 
  filter(str_detect(NAME, "^{{town_name}}"))

# Calculate total LEP households and percentage
lep_total <- town_lep_raw |> 
  filter(variable == "LEP Households") |> 
  pull(estimate)

hh_total <- town_lep_raw |> 
  filter(variable == "Total Households") |> 
  pull(estimate)

lep_percent <- if_else(hh_total > 0, round((lep_total / hh_total) * 100, 1), 0)

# 2. Top 5 Languages (C16001)
lang_vars <- c(
  "Spanish" = "C16001_003",
  "French, Haitian, or Cajun" = "C16001_006",
  "German or other West Germanic" = "C16001_009",
  "Russian, Polish, or other Slavic" = "C16001_012",
  "Other Indo-European" = "C16001_015",
  "Korean" = "C16001_018",
  "Chinese (incl. Mandarin, Cantonese)" = "C16001_021",
  "Vietnamese" = "C16001_024",
  "Tagalog (incl. Filipino)" = "C16001_027",
  "Other Asian and Pacific Island" = "C16001_030",
  "Arabic" = "C16001_033",
  "Other Languages" = "C16001_036"
)

town_lang_raw <- get_acs(
  geography = "county subdivision",
  variables = lang_vars,
  state = "MA",
  year = current_acs_year
) |> 
  filter(str_detect(NAME, "^{{town_name}}"))

# Sort and keep the top 5 spoken languages
top_5_languages <- town_lang_raw |> 
  arrange(desc(estimate)) |> 
  filter(estimate > 0) |>  # Exclude languages with zero speakers
  slice_head(n = 5) 
```

In **{{town_name}}**, there are **`r scales::comma(lep_total)`** households considered to have Limited English Proficiency (LEP), representing **`r lep_percent`%** of total households.
The top languages spoken at home (other than English) are:

```{r}
#| label: table-lang-{{town_name}}
#| echo: false

# Generate a simple, clean table
top_5_languages |> 
  select(Language = variable, Estimate = estimate) |> 
  knitr::kable(format = "markdown", format.args = list(big.mark = ","))
```


### Race & Ethnicity

```{r}
#| label: process-race-{{town_name}}
#| include: false

# 1. Race categories (total, regardless of Hispanic origin) — B02001
race_vars <- c(
  "White alone"                                       = "B02001_002",
  "Black or African American alone"                   = "B02001_003",
  "American Indian and Alaska Native alone"           = "B02001_004",
  "Asian alone"                                       = "B02001_005",
  "Native Hawaiian and Other Pacific Islander alone"  = "B02001_006",
  "Some other race alone"                             = "B02001_007",
  "Two or more races"                                 = "B02001_008"
)

# 2. Hispanic/Latino origin — B03003
hispanic_vars <- c(
  "Hispanic or Latino"     = "B03003_003",
  "Not Hispanic or Latino" = "B03003_002"
)

# Total population denominator
total_pop_var <- c("Total" = "B02001_001")

town_race_raw <- get_acs(
  geography = "county subdivision",
  variables = race_vars,
  state = "MA",
  year = current_acs_year,
  survey = "acs5"
) |>
  filter(str_detect(NAME, "^{{town_name}}"))

town_hispanic_raw <- get_acs(
  geography = "county subdivision",
  variables = hispanic_vars,
  state = "MA",
  year = current_acs_year,
  survey = "acs5"
) |>
  filter(str_detect(NAME, "^{{town_name}}"))

town_total_pop <- get_acs(
  geography = "county subdivision",
  variables = total_pop_var,
  state = "MA",
  year = current_acs_year,
  survey = "acs5"
) |>
  filter(str_detect(NAME, "^{{town_name}}")) |>
  pull(estimate)

# Build race data frame with a group label
town_race_clean <- town_race_raw |>
  filter(estimate > 0) |>
  mutate(
    variable = recode(variable,
      "White alone"                                       = "White",
      "Black or African American alone"                   = "Black / African American",
      "American Indian and Alaska Native alone"           = "American Indian / AK Native",
      "Asian alone"                                       = "Asian",
      "Native Hawaiian and Other Pacific Islander alone"  = "Nat Hawaiian / Pac Islander",
      "Some other race alone"                             = "Other race",
      "Two or more races"                                 = "Two or more races"
    ),
    share = round((estimate / town_total_pop) * 100, 1),
    group = "Race"
  )

# Build Hispanic origin data frame with a group label
town_hispanic_clean <- town_hispanic_raw |>
  mutate(
    variable = recode(variable,
      "Hispanic or Latino"     = "Hispanic / Latino",
      "Not Hispanic or Latino" = "Not Hispanic / Latino"
    ),
    share = round((estimate / town_total_pop) * 100, 1),
    group = "Hispanic or Latino Origin"
  )

# Combine, preserving group order
town_race_ethnicity <- bind_rows(town_race_clean, town_hispanic_clean) |>
  mutate(
    group = factor(group, levels = c("Race", "Hispanic or Latino Origin")),
    variable = reorder(variable, share)
  )
```

```{r}
#| label: prose-race-{{town_name}}
#| include: false

# Largest racial group
modal_race <- town_race_clean |>
  slice_max(estimate, n = 1, with_ties = FALSE) |>
  pull(variable)

modal_race_pct <- town_race_clean |>
  slice_max(estimate, n = 1, with_ties = FALSE) |>
  pull(share)

# Hispanic/Latino count and share
hispanic_count <- town_hispanic_clean |>
  filter(variable == "Hispanic / Latino") |>
  pull(estimate)

hispanic_pct <- town_hispanic_clean |>
  filter(variable == "Hispanic / Latino") |>
  pull(share)
```

In **{{town_name}}**, **`r modal_race`** is the largest racial group at **`r modal_race_pct`%** of the population. **`r scales::comma(hispanic_count)`** residents (**`r hispanic_pct`%**) identify as Hispanic or Latino, which is reported separately as it is an ethnicity rather than a racial category.

```{r}
#| label: plot-race-{{town_name}}
#| echo: false

ggplot(town_race_ethnicity, aes(x = variable, y = estimate, fill = group)) +
  geom_col(width = 0.7) +
  geom_text(
    aes(label = scales::comma(estimate)),
    hjust = -0.1,
    size = 2.5
  ) +
  scale_fill_manual(
    values = c("Race" = "#A00D03", "Hispanic or Latino Origin" = "#0395A0"),
    guide = "none"
  ) +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.35))
  ) +
  coord_flip() +
  facet_grid(group ~ ., scales = "free_y", space = "free_y") +
  labs(x = NULL, y = "Population") +
  theme(
    strip.text = element_blank(),
    strip.background = element_blank(),
    panel.spacing = unit(0.6, "lines")
  )
  
```

```{=typst}
#colbreak()
```

####
### Income & Poverty
```{r}
#| label: process-income-{{town_name}}
#| include: false

# 1. Household Income (B19001)
income_vars <- c(
  "Less than $10,000"      = "B19001_002",
  "$10,000 to $14,999"     = "B19001_003",
  "$15,000 to $19,999"     = "B19001_004",
  "$20,000 to $24,999"     = "B19001_005",
  "$25,000 to $29,999"     = "B19001_006",
  "$30,000 to $34,999"     = "B19001_007",
  "$35,000 to $39,999"     = "B19001_008",
  "$40,000 to $44,999"     = "B19001_009",
  "$45,000 to $49,999"     = "B19001_010",
  "$50,000 to $59,999"     = "B19001_011",
  "$60,000 to $74,999"     = "B19001_012",
  "$75,000 to $99,999"     = "B19001_013",
  "$100,000 to $124,999"   = "B19001_014",
  "$125,000 to $149,999"   = "B19001_015",
  "$150,000 to $199,999"   = "B19001_016",
  "$200,000 or more"       = "B19001_017"
)

income_total_var <- c("Total Households" = "B19001_001")

# 2. Poverty Status (B17001)
poverty_vars <- c(
  "Below poverty level"    = "B17001_002",
  "At or above poverty level" = "B17001_031"
)

poverty_total_var <- c("Total" = "B17001_001")

town_income_raw <- get_acs(
  geography = "county subdivision",
  variables = income_vars,
  state = "MA",
  year = current_acs_year,
  survey = "acs5"
) |>
  filter(str_detect(NAME, "^{{town_name}}"))

town_income_total <- get_acs(
  geography = "county subdivision",
  variables = income_total_var,
  state = "MA",
  year = current_acs_year,
  survey = "acs5"
) |>
  filter(str_detect(NAME, "^{{town_name}}")) |>
  pull(estimate)

town_poverty_raw <- get_acs(
  geography = "county subdivision",
  variables = poverty_vars,
  state = "MA",
  year = current_acs_year,
  survey = "acs5"
) |>
  filter(str_detect(NAME, "^{{town_name}}"))

town_poverty_total <- get_acs(
  geography = "county subdivision",
  variables = poverty_total_var,
  state = "MA",
  year = current_acs_year,
  survey = "acs5"
) |>
  filter(str_detect(NAME, "^{{town_name}}")) |>
  pull(estimate)

# Aggregate income into 5 brackets
town_income_clean <- town_income_raw |>
  mutate(
    bracket = case_when(
      variable %in% c("Less than $10,000", "$10,000 to $14,999",
                      "$15,000 to $19,999", "$20,000 to $24,999")           ~ "Under $25k",
      variable %in% c("$25,000 to $29,999", "$30,000 to $34,999",
                      "$35,000 to $39,999", "$40,000 to $44,999",
                      "$45,000 to $49,999")                                  ~ "$25k–$50k",
      variable %in% c("$50,000 to $59,999", "$60,000 to $74,999",
                      "$75,000 to $99,999")                                  ~ "$50k–$100k",
      variable %in% c("$100,000 to $124,999", "$125,000 to $149,999",
                      "$150,000 to $199,999")                                ~ "$100k–$200k",
      variable == "$200,000 or more"                                         ~ "$200k+"
    ),
    bracket = factor(bracket, levels = c("Under $25k", "$25k–$50k", "$50k–$100k",
                                          "$100k–$200k", "$200k+"))
  ) |>
  group_by(bracket) |>
  summarize(
    estimate = sum(estimate, na.rm = TRUE),
    moe = sqrt(sum(moe^2, na.rm = TRUE))
  ) |>
  mutate(
    share = round((estimate / town_income_total) * 100, 1),
    variable = as.character(bracket),
    group = "Household Income"
  )

# Poverty status
town_poverty_clean <- town_poverty_raw |>
  mutate(
    share = round((estimate / town_poverty_total) * 100, 1),
    group = "Poverty Status"
  )

# Combine
town_income_poverty <- bind_rows(
  town_income_clean |> select(variable, estimate, share, group),
  town_poverty_clean |> select(variable, estimate, share, group)
) |>
  mutate(
    group = factor(group, levels = c("Household Income", "Poverty Status")),
    variable = factor(variable, levels = c(
      "Under $25k", "$25k–$50k", "$50k–$100k", "$100k–$200k", "$200k+",
      "Below poverty level", "At or above poverty level"
    ))
  )


# Pull key figures for inline text
income_below_25k <- town_income_clean |>
  filter(bracket == "Under $25k") |>
  pull(estimate)

income_below_25k_pct <- town_income_clean |>
  filter(bracket == "Under $25k") |>
  pull(share)

income_top_bracket <- town_income_clean |>
  filter(bracket == "$200k+") |>
  pull(estimate)

income_top_bracket_pct <- town_income_clean |>
  filter(bracket == "$200k+") |>
  pull(share)

poverty_count <- town_poverty_clean |>
  filter(variable == "Below poverty level") |>
  pull(estimate)

poverty_pct <- town_poverty_clean |>
  filter(variable == "Below poverty level") |>
  pull(share)

# Find the modal (largest) income bracket for the narrative
modal_bracket <- town_income_clean |>
  slice_max(estimate, n = 1) |>
  pull(variable)

modal_bracket_pct <- town_income_clean |>
  slice_max(estimate, n = 1) |>
  pull(share)
```

In **{{town_name}}**, the largest share of households falls in the **`r modal_bracket`** income range, representing **`r modal_bracket_pct`%** of all households. **`r scales::comma(income_below_25k)`** households (**`r income_below_25k_pct`%**) report income under $25,000, while **`r scales::comma(income_top_bracket)`** households (**`r income_top_bracket_pct`%**) report income of $200,000 or more.
```{r}
#| label: plot-income-{{town_name}}
#| echo: false
#| fig-height: 2.5

ggplot(town_income_clean, aes(x = bracket, y = estimate)) +
  geom_col(fill = "#0395A0", width = 0.7) +
  geom_text(
    aes(label = paste0(scales::comma(estimate), "\n(", share, "%)")),
    hjust = -0.1,
    size = 2.2,
    lineheight = 0.9
  ) +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.35))
  ) +
  coord_flip() +
  labs(x = NULL, y = "Households")
```

**`r scales::comma(poverty_count)`** residents (**`r poverty_pct`%**) in **{{town_name}}** have incomes below the federal poverty level.
```{r}
#| label: plot-poverty-{{town_name}}
#| echo: false
#| fig-height: 1.5

ggplot(town_poverty_clean, aes(x = variable, y = estimate)) +
  geom_col(fill = "#A00D03", width = 0.7) +
  geom_text(
    aes(label = paste0(scales::comma(estimate), "\n(", share, "%)")),
    hjust = -0.1,
    size = 2.2,
    lineheight = 0.9
  ) +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.35))
  ) +
  coord_flip() +
  labs(x = NULL, y = "Population")
```

### Transportation to Work and Vehicles Available

```{r}
#| label: process-transportation-{{town_name}}
#| include: false

# 1. Vehicles Available (B08201)
veh_vars <- c(
 "Total Households" = "B08201_001",
  "No vehicle" = "B08201_002", 
 "1 vehicle" = "B08201_003",
 "2 vehicles" = "B08201_004",
 "3 vehicles" = "B08201_005", 
 "4 or more vehicles" = "B08201_006"
)

town_veh_raw <- get_acs(
  geography = "county subdivision",
  variables = veh_vars,
  state = "MA",
  year = current_acs_year
) |> 
  filter(str_detect(NAME, "^{{town_name}}"))

# Calculate Zero-Vehicle Households
zero_veh_total <- town_veh_raw |> 
  filter(variable == "No vehicle") |> 
  pull(estimate)

total_hh_veh <- town_veh_raw |> 
  filter(variable == "Total Households") |> 
  pull(estimate)

zero_veh_percent <- if_else(total_hh_veh > 0, round((zero_veh_total / total_hh_veh) * 100, 1), 0)

# 2. Means of Transportation to Work (B08301)
transpo_vars <- c(
  "Total Workers" = "B08301_001",
  "Drove alone" = "B08301_003", 
  "Carpooled" ="B08301_004", 
  "Public transportation" = "B08301_010", 
  "Taxi or ride-hailing services" = "B08301_016",
  "Motorcycle" = "B08301_017", 
  "Bicycle" = "B08301_018", 
  "Walked" = "B08301_019", 
  "Other means" = "B08301_020", 
  "Worked from home" = "B08301_021"
)

town_transpo_raw <- get_acs(
  geography = "county subdivision",
  variables = transpo_vars,
  state = "MA",
  year = current_acs_year
) |> 
  filter(str_detect(NAME, "^{{town_name}}"))

# Clean up for the chart
town_transpo_clean <- town_transpo_raw |> 
  filter(variable != "Total Workers", estimate > 0) |> 
  mutate(variable = reorder(variable, estimate))
```

In **{{town_name}}**, there are **`r scales::comma(zero_veh_total)`** households with no vehicles available, which represents **`r zero_veh_percent`%** of all households in the community.

**Means of Transportation to Work**

```{r}
#| label: plot-transpo-{{town_name}}
#| echo: false

ggplot(town_transpo_clean, aes(x = estimate, y = variable)) +
  geom_col(fill = "#0395A0", width = 0.7) +
  geom_text(aes(label = scales::comma(estimate)), hjust = -0.1, size = 2.5) +
  scale_x_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.2))) +
  labs(
    x = "Number of Workers",
    y = NULL
  )
```

```{=typst}
]
```

```{=typst}
#pagebreak()
```